# Working with simple data

### The command line

The command line is the primary mechanism that you'll use to interact with R.  When you enter instructions the R interpreter will perform computations for you. 

While this may seem like an arcane method for interacting with software, it has one *huge* advantage over point-and-click environments:  it is incredibly easy to repeat or abstract computations that you need to do often or on very large data sets.  Any instruction that you type at the R command line can also be saved to a "R script" file.  These files are just plain text files (by convention, R scripts have a ".r" or ".R" at the end of the file name).  Running an R script is identical to hand typing all of the commands in that script at the command line.

Let's try entering some simple math expressions to see how this interaction with the command line works.  In the code examples you'll see on this site the commands you type are followed by the text or visual output that R produces.

```{r}
1 + 2
2 * 3
4 ^ 5
6.7 / 8.9
````

As you might expect, R recognizes the standard syntax for numbers and mathematical operators.  When you've enter a complete [expression][] at the command line and hit "enter", R [evaluates](http://en.wikipedia.org/wiki/Evaluation_strategy) the result of that expression.  If you don't tell R what to do with the result, it will just print out a representation of the value on the next line.

Often, though, you will want to perform a calculation and save the result for later use.  If you've used Excel, you've probably used cells on a worksheet to hold the results of calculations based on the data in other cells.  In R we can save the results of computations in [variables][].  To do this we'll use the assignment operator, which is a little arrow in R `<-`.

(If you're coming from another programming language and the arrow syntax bugs you, you *can* use `=` operator for general assignment -- see the R help pages for a few important edge cases.)

```{r}
a <- 10
a
b <- a + 11
b
c <- a / b
c
````

If you're working with very large numbers there's a shorthand syntax you can use for scientific notation:

```{r}
2e10
2 * 10^10
2e10 == 2 * 10^10
````

In that last line we used the comparison operator `==`; it tests whether or not two values are equivalent.

Finally, let's see what happens if we try hitting `enter` before we've finished entering a complete expression.  Type `2 +` and then hit `enter`.  You'll see a little `+` shows up as your prompt on the next line.  This is R's way of telling you the text you've entered so far isn't a complete thought.  Finish this expression by entering another number and hitting `enter` again.

### Everything is a vector

In the above expressions we appeared to be doing computations on single numbers.  In fact, something more complicated was going on under the hood.  In R all data are actually [vectors](http://en.wikipedia.org/wiki/Array_data_structure) of data.  

Unlike other programming languages, there are no atomic values in R; the most basic data structure is a vector.  Single values are just vectors with one element.  This may seem odd at first, until you consider the key implication:  all operators and functions in R are built to handle vectors of data, not just single values.  This means anything that can be done with a single number can also be done with a vector of numbers.  This is a language that was clearly designed by statisticians! 

You can see how many elements a vector holds using the `length` function:

```{r}
length(10)
length(c)
````

Above, `length` is the first example we've seen of an R [function][].  In programming, functions are analogous to their mathematical counter parts: they take in one or more values and evaluate to a new value.  The R syntax for running, or "calling", a function is in the form of `functionName(value1, value2, ...)`.

The `length` function has reported that we've been working with vectors of 1 element so far.  To **c**ompose a vector with more than one element, we'll use the `c()` function:

```{r}
c(1,2,3,4)
d <- c(5,6,7,8)
d + 10
````

As you can see, mathematical operators in R are built to handle vectorized operations:  we could add a 4 element vector `d` to a one element vector `10` and get a sensible result.

**Side note for programmers**:  if you're coming to R with a background in other programming languages, you might have caught something in the last code block that freaked you out.  We had previously assigned the variable `c` to hold a number; yet we were still able to call the built-in `c` function from this scope.  In almost any other common scripting language this would not have worked.  Although base scoping rules in R are strictly lexical, there are a number of aggressive additional checks that are performed on lookup or evalution failures.  For example, when an attempt is made to apply a non-function in an inner scope, the interpreter will ascend the scope chain looking for upvalues that *are* functions.  This is why we were able to bind `c` to a vector of numbers in our current scope and also apply the base R `c()` function.  Another example of this design philosophy at play in R is aggressive partial matching for named arguments in function calls.


### Strings and Boolean values

So far we've just been working with numbers (`numeric` values in R lingo), but R also supports text (`string` values) and true/false (`boolean` values) data.

To create strings, surround your text with either double `" ... "` or single `' ... '` quotes:

```{r}
"a"
'a'
"a" == 'a'
c( "a", "b", "c", "d" )
````

You can create strings containing single quotes using double quotes and visa versa, but if you need to make a string that contains both single and double quotes you need to use the `\` "escape" character:

```{r}
s <- "My data are \"awesome\"!"
s
message(s)
````

The two other special string characters that you're likely to run across when working with data files are the tab `\t` and newline `\n`:

```{r}
s <- "a\tb\tc"
s
message(s)
s <- "a\nb\nc"
s
message(s)
````

To create boolean values use `TRUE` or `FALSE`:

```{r}
TRUE
FALSE
TRUE == FALSE
c(TRUE, FALSE, TRUE, FALSE)
````


### Missing Data

R has built-in support for flagging values as missing data.  The `NA` value can be mixed with any kind of vector:

```{r}
c(1,2,NA,4)
c( "a", NA, "c", NA )
c(TRUE, FALSE, NA, FALSE)
````

Most R functions will either understand how to deal with missing data, or issue an error if they involve a type of statistical analysis that can't be used with missing data.

**A note about NULL**:  R also has a special "no-value" type called `NULL`.  If you are coming to R from another programming languages it is easy confuse the two (for example, in Python the `None` type is often used to do double duty, signifying `NULL` or `NA` in depending on the context).  By convention, you should use `NA` in data structures to represent missing data points (data that is *really* data even though it is a missing observation).  `NULL` should be used to refer to unassigned variables or unspecified arguments to functions.


### Getting help

Manual pages for all of the functions in base R and R packages are built into the language.  To quickly get help for any function you can use the `?` syntax.  For example, to get the help page for the `c()` function we used above enter `?c`.  RStudio also has a search box on the help tab.

[function]: http://en.wikipedia.org/wiki/Function_(programming)
[variables]: http://en.wikipedia.org/wiki/Variable_(computer_science)
[expression]: http://en.wikipedia.org/wiki/Expression_(computer_science)

