# Working with simple data


### The command line

The command line, or console, is the primary mechanism that you'll use to interact with R:  you enter instructions and the R interpreter performs computations for you.  If you have a common set of tasks that you need to repeat often, or a set of instructions that are too complicated to keep track of *ad hoc*, you can save a set of commands into a text file called an R script (by convention, these files have a ".r" or ".R" at the end).

Let's try entering some simple math expressions:

```{r }
1 + 2
2 * 3
4 ^ 5
6.7 / 8.9
````

As you might expect, R recognizes standard syntax for numbers and mathematical operators.  When you've entered a complete expression at the command line and hit "enter", R will evaluate the result of that expression.  If you don't tell R what to do with the result, it will just print out a representation of the value on the next line.

Often though, you want to perform a calculation and save the result for later use.  If you've used Excel, you've probably used cells on a worksheet to hold the results of calculations based on the data in other cells.  In R, we can save the results of computations in variables.  To do this, we'll use the assignment operator, which is a little arrow in R `<-` (if you're comming from another programming language and this syntax bothers you, you can use `=` for assignment -- see help pages for additional information about edge cases):

```{r}
a <- 10
a

b <- a + 11
b

c <- a / b
c
````

If you're working with very large numbers, common in the sciences, there's a shorthand syntax you can use:

```{r}
2e10
2 * 10^10
2e10 == 2 * 10^10
````

In that last line we used the comparison operator `==`; it tests whether or not two values are equivalent.


Finally, let's see what happens if we try hitting `enter` before we've finished entering a complete expression.  Type `2 +` and then hit `enter`.  You'll see a little `+` shows up as your prompt on the next line.  This is R's way of telling you the text you've entered so far isn't a complete thought.  Finish this expression by entering another number and hitting `enter` again.


### Everything is a vector

In the above expressions we appeared to be doing computations on single numbers.  In fact, something more complicated was going on under the hood.  In R all data are actually vectors of data.  Unlike other programming languages, there are no atomic values; the most basic data structure is a vector.  This may seem odd until you consider the key implication:  all operators and functions in R are built to operate on vectors of data, not just single values.  Single values are just vectors with one element.

You can see how many elements a vector holds using the `length` function:

```{r}
length(10)
length(c)
````

Above, `length` is the first example we've seen of an R function.  In programming functions are analogous to their mathematical counter parts: the take in one or more values and evaluate to a possibly new value.  The syntax for running, or "calling", a function will always be in the form of `functionName(value1, value2, ...)`.

The `length` function has reported that we've been working with vectors of 1 element so far.  To **c**ompose a vector with more than one element, we'll use the `c()` function:

```{r}
c(1,2,3,4)
d <- c(5,6,7,8)
d + 10
````

As you can see, mathematical operators in R are built to handle vectorized operations:  we could add a 4 element vector `d` to a one element vector `10` and get a sensible result.

**Side note for programmers**:  if you're comming to R with a background in other programming languages, you might have caught something in the last code block that freaked you out.  We had previously assigned the variable `c` to hold a number; yet we were still able to call the built-in `c` function from this scope.  In almost any other common scripting language this would not have worked.  Although base scoping rules in R are strictly lexical, there are a number of aggressive additional checks done on failures.  For example, when an attempt is made to apply a non-function in an inner scope, the interpreter will check up-values.  Another example of this design philosophy at play in R is aggressive partial matching for named arguments in function calls.


### Strings and Boolean values

So far we've just been working with numbers (`numeric` values in R lingo), but R also supports text (`string` values) and true/false (`boolean` values) data.

To create strings, surround your text with either double `" ... "` or single `' ... '` quotes:

```{r}
"a"
'a'
"a" == 'a'
c( "a", "b", "c", "d" )
````

You can create strings containing single quotes using double quotes and visa versa, but if you need to make a string that contains both single and double quotes you need to use the `\` "escape" character:

```{r}
s <- "My data are \"awesome\"!"
s

message(s)
````

The two other special string characters that you're likely to run across when working with data files are the tab `\t` and newline `\n`:

```{r}
s <- "a\tb\tc"
s
message(s)

s <- "a\nb\nc"
s
message(s)
````

To create boolean values use `TRUE` or `FALSE`:

```{r}
TRUE
FALSE
TRUE == FALSE
c(TRUE, FALSE, TRUE, FALSE)
````


### Missing Data

R has built-in support for flagging values as missing data.  The `NA` can be mixed with any kind of vector:

```{r}
c(1,2,NA,4)
c( "a", NA, "c", NA )
c(TRUE, FALSE, NA, FALSE)
````

Most R functions will either understand how to deal with missing data, or issue an error if they involve a type of statistical analysis that can't be used with missing data.

**A note about NULL**:  R also has a special "no-value" type called `NULL`.  Users comming from other programming languages often confuse the two (Python's None type, for example, does double duty).  By convention, you should use `NA` in data structures to represent missing data points; `NULL` should be used to refer to unassigned variables or unspecified arguments to functions.


### Getting help

Manual pages for all of the functions in base R and R packages are built into the language.  To quickly get help for any function you can use the `?` syntax.  For example, to get the help page for the `c()` function we used above enter `?c`.  RStudio also has a search box on the help tab.

