# Working with Tables

R has great built-in support for working with data in tabular format.  Tables in R are called "data frames".  By convention, response and annotation variables are arranged across the columns and observations down the rows.  Columns, and optionally rows, can also be given unique name.



### blag

## Look at the contents of translationTable.txt

This is a "tab-delimited" file.

You can make one in a text editor, Excel, Minitab, etc...

## The read.* family of functions load tabular data into R

``` {r}
#Read a tab-delimited file
translationTable <- read.delim("translationTable.txt", stringsAsFactors = FALSE)
````

What's up with the `stringsAsFactors = FALSE` part?  Factors are a more advanced topic, so we're not going to get into them today (but we *will*).  Try ommiting this argument to `read.delim` and see what happens when you run through the exercises below.  It's close to the same result.

## If you have comma-separated file

``` {r error=FALSE, warning=FALSE}
myTable <- read.csv("myTable.csv")
````

## If you need finer control over how the table is loaded

``` {r}
?read.table
````

# Tables

## It's not a number, it's not a string...

``` {r}
#Class tells you what kind of data a variable holds

#So far we've seen
class("hello")
class(100)

#We've also used TRUE as an argument for a function...
class(TRUE)

#The other logical value is FALSE
class(FALSE)

#So what's our translationTable?
class(translationTable)
````

## The R terminology for a table is a "data.frame"

How does the structure of translationTable related to our original file?

  row.names |  aminoacid
  --------- | ---------
  GCU	      | A
  GCC	      | A
  GCA	      | A

The relationship between the column headings in a data.frame the the original file can be a bit tricky to understand when you first start working with tables in R.

## Two flavors of data.frame (depending on your data)

* Unique values in the 1st column (row.names)
* Non-unique values in the 1st column (no row.names)

When you use one of the simple read.* functions, R will try to figure out whether or not it can treat the values in the left-most column as unique.  If it can, it will make a `data.frame` with a `row.names` column on the left side of the table.  In this case, you want the heading line of your file to have one *fewer* labels than you have columns.  See how I did this in translationTable.txt?

If your left most column does **not** contain unique values, R will add a new left-hand column that just contains numbers (1..n where n is the number of rows you have).  You'll see why we want to have some kind of unique label for each row in our `data.frame` objects in a moment.

# Sidebar:  Extract & Replace

Remember how we mentioned that the fundamental data type in R is the vector and then I promised that this would be both hard for you to wrap your heads around at first, but also extremely powerful?  Well, prepare to have your minds blown because we're about to see our first real `vectorized` operations in R!

## Let's say we have a vector of numbers:
``` {r}
myNumbers <- c(10,20,30,40,50)
````

Here's how we can `extract` a subset of vector:
``` {r}
myNumbers
myNumbers[1]
myNumbers[3]
````

## What if we want more than one at once?
``` {r}
myNumbers
myNumbers[c(1,3)]
````

What happened there?

## How can we extract a range of numbers?
``` {r}
# The : operator creates a series of numbers
1:5
2:10

# We can use this inside of our [...]
myNumbers
myNumbers[2:3]
````

## We can also use logical vectors

TRUE = keep it
FALSE = don't

``` {r}
myNumbers
myNumbers[c(FALSE,TRUE,TRUE,TRUE,TRUE)]
myNumbers[c(TRUE,FALSE,FALSE,FALSE,FALSE)]
````

## Now things get interesting!

``` {r}
myNumbers
myNumbers[TRUE]
myNumbers[c(TRUE,FALSE)]
````

## All operations in R are vectorized...

Yesterday we saw that the `+` works on vectors. 

Others basic math operators:
``` {r warning=FALSE}
myNumbers
myNumbers + 5
myNumbers - 5
myNumbers * c(10,0)
myNumbers / 0.1
````

## ...including comparison!

``` {r}
myNumbers
myNumbers > 25
myNumbers < 25

#Equality comparison uses == instead of =
myNumbers == 30

# The `!` operator negates logical vectors ("not")
myNumbers != 30
!(myNumbers > 25)
````

To compare equality we use the `==` operator.  This is to prevent confusion with the `=` which is used to pass arguments to functions (or do assignment like `<-`).  This will probably trip you up at first, but you'll get used to it.  See what happens if you run `myNumbers = 30`.


## So what does it all mean?

We can use our extraction operators `[` `]` to filter data in our vectors.

This is a very common idiom in R:
``` {r}
myNumbers
myNumbers[myNumbers > 25]
myNumbers[myNumbers < 25]
````

## You can get fancy...

Explore the parts of this code, then explain what it does:
``` {r}
myNumbers[ (myNumbers %% 4) == 0 ]
````

## You can use [] to replace data too!
``` {r}
myNumbers
myNumbers[3] <- 100
myNumbers
myNumbers[2:3] <- c(1,2)
myNumbers
````

# Back to our table

## The [] operators work with data.frames as well

``` {r}
# You can subset tables on a column: table[col]
## translationTable[1]

# You can subset tables on both a row and a column
translationTable[3,1]

#If you leave out the column, but keep the , you'll get all columns
translationTable[3,]
````

## You can use text to refer to columns or rows by name

Only works for rows if your `data.frame` has unique `row.name`'s

``` {r}
translationTable["GUC","aminoacid"]
translationTable["GUC",]

#Oh hey!  Remember vectors?
translationTable[c("GUA","UAG"),]

#... and logicals?
translationTable[c(TRUE,FALSE),]
````

## A couple other tricks...
``` {r}
#If your table has row names, you can get a vector of them with:
row.names(translationTable)

#The $ is a short hand notation for selecting a column of data from a table
translationTable$aminoacid
````

## Your turn #1

Subset the table to return a vector of all of the codons which encode valine
"V"

## Your turn #2

Write a function that takes a RNA codon and returns the amino acid.

``` {r}
translateRNA <- function(codon) {

  #Insert your code here

}
````

``` {r include=FALSE}
translateRNA <- function(codon) {
  translationTable[codon,]
}
````

Make sure it works:
``` {r}
translateRNA("UAU")
````

How might this function be useful in a script designed to translate dna.txt?

For a full reference on extraction and replacement operators in R, you can search for "[" in the Rstudio search box, or run `help("[")` at the prompt.

# Subsetting strings

## We're getting close!  

We can translate codons...
``` {r}
translateRNA("UGG")
````
...but we need to be able to read a string of letters in sets of 3.
``` {r}
sampleRNA <- "GTAGCTGGCAAGTTAATAGAGATGGAAAAC"
````
